---
title: "The _Quickly_ Word Prediction App"
author: "Carl Turner"
date: "April 20, 2016"
output: html_document
---

##Background
Presenting mobile users with predictions of text can significantly improve the usability of entering text on small keyboards. This project attempts to balance speed and accuracy in a text prediction model by 1) using small samples of a large text database and 2) creating a fast algorithm for predicting next word text from user input. This approach sacrifices some accuracy for speed. 

##Initial Line and Word Count
This first step simply opens the US English blog file, reads the file, and outputs it as an RDS file to be used in the Shiny app. 

```{r, cache=FALSE, label="read file"}
library(ngram)
strTime <- Sys.time()
fileName <- "./data/final/en_US/en_US.blogs.txt"
con <- file(fileName,open="r")
txt <- readLines(con, skipNul = TRUE, n=3000)
close(con)

saveRDS(txt, "shiny/data/blogtxt.rds") #used in server.R
txt<-readRDS("shiny/data/blogtxt.rds") #used in server.R
endTime <- Sys.time()
timeRun <- difftime(endTime,strTime,units="auto")
timeRun
```

##Ngrams of Data Source
In this section the frequencies for individual words, 2-grams (bigrams), and 3-grams (trigrams), are caluculated for US blogs.

###Single Letter
Pre-processing is done by removing non-alpha characters before computing word frequencies. Then the words are sorted by their frequency of occurance. The 10 most-frequent words are shown below.

I found that memory limitations of my own machine prevented me from scanning the entire news file. 

```{r, cache=TRUE, label="blog single letter"}
library(ngram)
library(stringr)
blogtxt = scan(text = txt, what="", quote=NULL, blank.lines.skip = TRUE)#######

blogtxt<-gsub("[^[:alnum:][:space:]\']", " ", blogtxt) #remove all punctuation except "'"
blogtxt<-gsub("â ", "\'", blogtxt)
blogtxt<-gsub("\\' ", "\\'", blogtxt)
blogtxt<-gsub(" [b-hj-z] "," ", blogtxt) #remove single letters except a,i
blogtxt<-preprocess(blogtxt)
blogtxt<-preprocess(blogtxt, case="lower", split.at.punct=FALSE)
blogcounts = as.data.frame(xtabs(~blogtxt))
blog.str = paste(blogtxt, collapse=" ")

library(tau)
blog.counts = textcnt(blog.str, n=1, method="string", tolower=T) #1-gram
blog.counts.df = data.frame(word = names(blog.counts), count = c(blog.counts))
head(blog.counts.df[order(blog.counts.df$count, decreasing = T),], 10)
sum(blog.counts.df$count) #word count
```

###2-grams
The 2-grams (bigrams) are created by taking each pair of words, then calculating the frequencies that each pair appears in the corpus. As shown in the figure, there are a small number of high-frequency 2-grams. The high-frequency 2-grams are shown in the table below the figure. 

```{r, cache=TRUE, label="blog 2gram"}
library(ggplot2)
library(NLP)
library(tm)
blog.corpus = Corpus(VectorSource(blog.str))
blog.corpus = tm_map(blog.corpus, tolower) 
cleaned.blog.str = as.character(blog.corpus)[1]
blog.words = strsplit(cleaned.blog.str, " ", fixed = T)[[1]]

blog.bigrams = vapply(ngrams(blog.words, 2), paste, "", collapse = " ")
blog.bigram.counts = as.data.frame(xtabs(~blog.bigrams))
blog.bigrams.y <- blog.bigram.counts[order(blog.bigram.counts$Freq, decreasing = T),]
y<-blog.bigrams.y
x<-c(1:length(y$Freq))

qplot(x, y$Freq, 
      main = "US Blogs", 
      xlab = "ngram = 2",
      ylab = "frequency", 
      xlim = c(0, length(y$Freq)/10),
      na.rm = TRUE,
      geom = c("line", "point"))
head(y)

```

###3-grams
The 3-grams (trigrams) are created from trios of words, then frequencies are calculated for each 3-gram. The distribution of 3-grams are shown in the figure and the names of the most frequent 3-grams are in the table below the figure. 

```{r, cache=TRUE, label="blog 3gram"}
library(ggplot2)
library(NLP)
library(tm)
blog.trigrams = vapply(ngrams(blog.words, 3), paste, "", collapse = " ")
blog.trigram.counts = as.data.frame(xtabs(~blog.trigrams))
blog.trigrams.y <- blog.trigram.counts[order(blog.trigram.counts$Freq, decreasing = T),]
y<-blog.trigrams.y
x<-c(1:length(y$Freq))

qplot(x, y$Freq, 
      main = "US Blogs", 
      xlab = "ngram = 3",
      ylab = "frequency", 
      xlim = c(0, length(y$Freq)/10),
      na.rm = TRUE,
      geom = c("line", "point"))
head(y)
```

##Prediction Function

###Matching Routine

```{r, cache=FALSE, label="matching routine"}
library(stringr)
trimInput <- function(userinp) {
    userinp<-str_replace_all(userinp, fixed("  "), " ")
    userinp<-str_trim(userinp) 
    if (length(unlist(strsplit(userinp, " "))) < 3){
        userinp
    } else {
        userinpLen <- length(unlist(strsplit(userinp," ")))
        paste(word(userinp, userinpLen-1), word(userinp, userinpLen))
    }
}
getTrigram <- function(matches, userinp) {
    for(i in 1:length(matches$blog.trigrams)){
        if (word(matches$blog.trigrams[i],1) == word(userinp, 1)){
            #print(matches$blog.trigrams[i])
            return(word(matches$blog.trigrams[i], 3))
            break
        }
    } 
}
getBigram <- function(matches, userinp) {
    for(i in 1:length(matches$blog.bigrams)){
        if (word(matches$blog.bigrams[i],1) == word(userinp, 2)){
            #print(matches$blog.bigrams[i])
            return(word(matches$blog.bigrams[i], 2))
            break
        }
    } 
}
getPrediction <- function(userinp) {
    userinp <- trimInput(userinp)
    matches<-blog.trigrams.y[grep(userinp, blog.trigrams.y$blog.trigrams), ]
    if (dim(matches)[1]>0){
        getTrigram(matches, userinp)
    } else {
        matches<-blog.bigrams.y[grep(word(userinp, 2), blog.bigrams.y$blog.bigrams), ]
        if (dim(matches)[1]>0){
            getBigram(matches, userinp)
        } else {
            return("no prediction")
        }
    }
}

```

###Test Cases
```{r, label="test cases"}
userinp <- c("one of") #a top trigram returns 'the'
#userinp <- c("one in") #not in trigram, but in bigram. "of" returns 'the'
#userinp <- c("one only") #not in trigram or bigram
getPrediction(userinp)
```
